Covariate analysis of CMC RNASeq data reprocessed with Star-HTSeq pipeline (DLPFC with TMM and SVA normlaisation)

### 2016-09-19 code/Rmd/CMC_DLPFC_StarAligned_TMM_SVA.Rmd
Added markdown to process star aligned DLPFC data with TMM and SVA based normalisation

---
title: "Covariate analysis of CMC RNASeq data reprocessed with Star-HTSeq pipeline (DLPFC with TMM normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knitr)
library(knit2synapse)
library(synapseClient)

@@

knit2synapse::knitToFolderEntity(file = "./CMC_DLPFC_StarAligned_TMM.Rmd",
                                 parentId ="syn5569976",
                                 entityName = 'DLPFC - STAR Reprocessed - with TMM')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5569976';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of star aligned CMC reprocessed RNASeq data with TMM normalisation (DLPFC)';

thisFileName <- 'CMC_DLPFC_StarAligned_TMM.Rmd'

@@

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. Here "Dx" is chosen as the primary variable of interest
```{r iterativeNorm}
# Primary variable of interest
postAdjustCovars = 'RIN';#'Dx'

# Assign residual covariates
residualSigCovars = preAdjustedSigCovars
covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]
postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()

loopCount = 0 
while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
  writeLines(paste('Using following covariates in the model:',
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects'))

  # Add postadjusted covariates (if any)
  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, 
                                                                          c(postAdjustCovars))]

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}

# Add primary variable to the list (in case not chosen by iterative method)
postAdjustCovars = unique(c('Dx', postAdjustCovars))

@@

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`
### Sanity check
```{r residualSigCovars.manual, fig.width=20, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Residual calculation
Calculate weighted residuals and add back "Dx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

@@

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC)) + geom_point() + xlim(c(-1,1))
p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
p = p + geom_vline(xintercept = -log2(1.2), color = 'red') 
p = p + facet_grid(.~Comparison, scales = 'fixed')
p
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "DLPFC - STAR Reprocessed - with TMM", parentId = parentId)
CODE <- synStore(CODE)

>>>>

---
title: "Covariate analysis of CMC RNASeq data reprocessed with Star-HTSeq pipeline (DLPFC with TMM and SVA normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knitr)
library(knit2synapse)
library(synapseClient)

@@

knit2synapse::knitToFolderEntity(file = "./CMC_DLPFC_StarAligned_TMM_SVA.Rmd",
                                 parentId ="syn5569976",
                                 entityName = 'DLPFC - STAR Reprocessed - with TMM and SVA')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5569976';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of star aligned CMC reprocessed RNASeq data with TMM and SVA normalisation (DLPFC)';

thisFileName <- 'CMC_DLPFC_StarAligned_TMM_SVA.Rmd'

@@

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. Here "Dx" is chosen as the primary variable of interest
```{r iterativeNorm}
# Primary variable of interest
postAdjustCovars = c('Dx', 'RIN', 'RIN2', 'Age_of_Death', 'PMI', 'IsolationBatch')

@@

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`
### Sanity check
```{r residualSigCovars.manual, fig.width=20, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Estimate surrogate variables using SVA package
```{r sva}
MODEL0 = DM1$design[,-grep('Dx',colnames(DM1$design))] # Get null model by removing variable of interest
MODEL1 = DM1$design

## Calculate the number of surrogate variables
NUM.SV = num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456, B = 10)

SURR.VAR = sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0, method = "irw", n.sv = NUM.SV, B = 30)$sv
colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])

DM1$design = cbind(MODEL1, SURR.VAR)

# Estimate correlation of random effects
VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F)

# Re-calculate linear fit using new weights and new design
VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
```
### Residual calculation
Calculate weighted residuals and add back "Dx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

@@

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC)) + geom_point() + xlim(c(-1,1))
p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
p = p + geom_vline(xintercept = -log2(1.2), color = 'red') 
p = p + facet_grid(.~Comparison, scales = 'fixed')
p
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "DLPFC - STAR Reprocessed - with TMM and SVA", parentId = parentId)
CODE <- synStore(CODE)
